Why there are more species in several small patches vs few large patches:

### 2024-09-25 ms_SLOSS.Rmd
Updated to submit revision 1

---
title: "Why there are more species in several small patches vs few large patches:
  a multi-species modeling approach"
bibliography: sloss_bib.bib
csl: "functional-ecology.csl"
output:
  pdf_document:
    includes:
      in_header: margins.sty
  word_document: default
---

@@

2. This work aims to explore spatial multi-species models to verify the SS > SL pattern and determine its causes. We used three types of models: a classical neutral, neutral with habitat selection, and a competitive hierarchical model.  All models have three parameters, namely extinction, reproduction, and dispersal distance. In the classical neutral model, species are equivalent and have the same parameters; when there is a reproduction event they send the propagule to a surrounding patch blindly, and if the patch is already occupied the propagule dies. In the neutral-with-habitat selection model, the propagules select an empty patch so their survival is ensured. In the hierarchical model, species have competitive dominance and more dominant ones can replace the less dominant with a replacement rate parameter. We use 100 species and start simulating the colonization of an empty area then we destroy a fixed amount of habitat using different regular patch sizes.

@@

    Regular degradation: Habitat is removed so that habitable patches form a regular pattern of squares with sides given by a habitat-size parameter. As the percentage of degraded habitat is constant varying the habitat-size parameter results in either a few large habitat patches (SL) or several small (SS).

    Random block degradation: Habitat square patches of a fixed size (habitat-size parameter) are located at random throughout the landscape. The sizes are identical to those used for regular degradation simulations, but it results in habitat patches that have a minimum size corresponding to the habitat-size but a maximum size that is bigger because habitat patches can overlap between them producing a larger patch. Then, we characterize this with the maximum habitat patch size.

@@

The parameters encompass a range of $growth-rates$ while holding the $death-rate$ fixed at 1. In this model, the relationship between these parameters, $\lambda = growth-rate / death-rate$, is the main determinant of the fate of an individual of a species. In similar non-spatial models, this relationship must exceed 1 for a species to persist. Thus we used a minimum $\lambda$ of 1.7, we also fixed the $migration-rate$ at 0.0001. We simulate each model variant using the parameters shown in Table 1. For the neutral classical version only, we also simulate random block habitat patterns. For each combination of parameters we ran 30 simulations up to 1500 time steps and recorded the number of species, the Shannon Diversity index [@Magurran2013] and the total number of individuals. For some representative combinations of parameters, we take snapshots of the model output showing the typical dynamics on time of the 3 recorded quantities and the rank abundance species distribution [@Newman2005; @May1975]

@@

In our neutral model simulations with a migration rate of $0.0001$ and regular habitat patterns, species richness varied with habitat patch sizes, influenced by dispersal distance and the $\lambda$ parameter (Figure 1). Species richness increased with $\lambda$, with the highest richness at smaller patch sizes for higher $\lambda$ values. Conversely, a $\lambda$ of 3 or less showed reduced richness at the smallest patch sizes for lower dispersal distances. In these cases, only migrants survive in the smallest patches, and no local populations are established (Figures S13 & S14).
The interplay of dispersal distances, habitat sizes, and $\lambda$ impacts community survival and species richness. For $\lambda < 4$ this suggests a threshold patch size influenced by these 3 parameters, for $\lambda = 4$ patch size decreases species richness. The differences between increasing $\lambda$'s almost disappear when dispersal distance is higher. Extended dispersal distances could enhance species survival in scenarios with low $\lambda$ and smaller patches but reduce richness beyond a certain threshold. Similar patterns are evident for Shannon diversity (Figure 2).

![Relationship between the number of species and habitat patch size for varying dispersal distances and $\lambda$ parameter (across panels). Simulations were performed using a neutral model with a constant migration rate (0.0001). Following 200 time steps of the model, 60% of the habitat was destroyed, rendering it unavailable for the species. The destruction of habitat resulted in the formation of regular patterns, where all remaining patches exhibited equal sizes.](Figures/Species_Habitat_Dispersal_Lambda_regular.png)

![Relationship between Shannon Diversity and habitat patch size for varying dispersal distances and $\lambda$ parameter (across panels). Simulations were performed using a neutral model with a constant migration rate (0.0001). Following 200 time steps of the model, 60% of the habitat was destroyed, rendering it unavailable for the species. The destruction of habitat resulted in the formation of regular patterns, where all remaining patches exhibited equal sizes.](Figures/Shannon_Habitat_Dispersal_Lambda_regular.png)

For the simulations conducted with randomly arranged block habitat patterns, the patches exhibit variable sizes, following a distribution with a minimum size determined by the patch-size parameter and a larger maximum size determined by the superposition of several patches (Figures S9 - S11). The overall pattern is analogous to the regular case, with a peak in richness occurring for intermediate patch sizes. However, for values of $\lambda$ higher than 2, the maximum richness is observed at the smaller habitat patch size (Figure S1 & S2). The patches in the random block simulations are generally larger, resulting in lower species richness and diversity, except for the minimum patch size (Figure S7 & S8). Supplementary material (Figures S9 & S32) includes examples of simulations with random block and regular habitat patterns. In these figures, it is evident that for the minimum patch size, only migrants survive in regular habitat patterns with neutral and hierarchical models.

![Relationship between the number of species and habitat patch size for different versions of the model (across panels) and varying dispersal distances. The neutral is the classical neutral model, species are equivalent and have the same parameters; in the hierarchical model, species have competitive dominance and more dominant ones can replace the less dominant; the neutral-with-habitat-selection model, is a neutral model but propagules search an empty patch so their survival is ensured. Simulations were performed using a fixed growth rate (2), migration and replacement rate of 0.0001 and 0.3 respectively. Following 200 time steps of the model, 60% of the habitat was destroyed, rendering it unavailable for the species. The destruction of habitat resulted in the formation of regular patterns, where all remaining patches exhibited equal sizes.](Figures/AcrossBehav_Species_Dispersal_Lambda2_regular.png)

![Relationship between Shannon Diversity and habitat patch size for different versions of the model (across panels) and varying dispersal distances. The neutral is the classical neutral model, species are equivalent and have the same parameters; in the hierarchical model, species have competitive dominance and more dominant ones can replace the less dominant; the neutral-with-habitat-selection model, is a neutral model but propagules search an empty patch so their survival is ensured. Simulations were performed using a fixed growth rate (2), migration and replacement rate of 0.0001 and 0.3 respectively. Following 200 time steps of the model, 60% of the habitat was destroyed, rendering it unavailable for the species. The destruction of habitat resulted in the formation of regular patterns, where all remaining patches exhibited equal sizes.](Figures/AcrossBehav_Shannon_Dispersal_Lambda2_regular.png)

Comparing the three versions of the model at $\lambda =2$, for higher dispersal the neutral model tends to have higher richness and Shannon diversity at higher patch sizes, followed by the hierarchical model and the neutral-with-habitat-selection model (Figures 3 & 4). At the lowest patch size, only in the neutral-with-habitat-selection have populations that can survive, because in the other model versions the individuals observed are migrants that cannot establish local populations and become extinct (See Figures S5 - S15). Thus an intermediate patch size with a maximum species richness is not observed for this model. Is also clearly observed that for the hierarchical model, there is a strong dominance of the more competitive species where Shannon Diversity is very low for a higher dispersal rate (Figure 4 & S23).

Examples of the spatial patterns of species and habitat distributions are illustrated in Figures 5 and S9-S32. Simulations with the neutral model depict the species' spatial distribution before fragmentation (Figure 5 A & B) and after fragmentation (Figure 5 C to F), with habitat destruction represented in purple areas. Immediately after fragmentation, the species' spatial distributions closely resemble those before habitat destruction, but then the patch compositions diverge. The divergence in patch species composition is more pronounced in the hierarchical model, where small patches tend to be dominated by one species (Figures S17 - S24). The effect of the minimum patch size for survival is also observed (Figure 5 E).

![Examples of species spatial patterns before and after fragmentation for the neutral model are illustrated. Different shades from dark green to white represent distinct species, with the destroyed habitat depicted in purple. Simulations were conducted with a fixed growth rate ($\lambda=2$), a migration rate of 0.0001, and a dispersal distance of 1.5. A. & B. showcase spatial patterns before fragmentation. C. & D. depict spatial patterns after fragmentation, where 60% of the habitat was destroyed, with patch sizes of 3 (C) and 29 (D). In these cases, the significance of the species' spatial distribution before fragmentation becomes evident. E. & F. display spatial patterns after 600 time steps, illustrating the extinction of populations at smaller patch sizes (E).](Figures/ModelSpatialPatternComparison.png)

@@

Our research contributes to the ongoing SLOSS debate in conservation biology, offering new insights based on our model findings. Consistent with previous studies, the results from our simulations primarily support the SS > SL pattern for species retention, as observed across various model scenarios [@Fahrig2020]. This trend underlines the greater efficacy of multiple smaller habitats in preserving species diversity compared to a single large habitat under equal conditions of habitat loss. A critical factor in this dynamic is the balance between reproduction and extinction rates ($\lambda$ parameter) and dispersal relative to the size of the habitat patches, influencing their role as sources of propagules. Notably, the SL > SS scenario emerges only when dispersal limitations prevent patch connectivity, and patch sizes are above a certain threshold to prevent local extinctions, aligning with Rybicki's [-@Rybicki2020] findings.

In neutral models, where species are considered equivalent, one species, by chance, tends to dominate a patch (Hubbell, 2001). Consequently, in fragmented landscapes, once a species increases its density, the likelihood of another species replacing it is low. With more patches, the probability of having a greater number of species is heightened. This effect is further amplified when individuals can actively select their habitat, leading to increased survival of propagules. Consequently, fragmentation exhibits a positive influence on the number of species in such scenarios.

@@

Edge effects are present in the neutral and hierarchical models, but not in the neutral-with-habitat-selection model. In the first two models, the edge effect is because individuals in the patch's border have fewer neighbors and the probability of extinction is higher.  In the neutral-with-habitat-selection model, the edge effect is not present because the individuals search for empty patches, so the individuals on the border of the patch have the same probability of extinction as the individuals in the interior of the patch. Thus when patches are smaller the edge proportion is higher and the edge effect is more important, but this plays a more important role in the minimal patch size for the survival of species than for the species richness.

The spatial arrangement of habitat patches, contrary to some expectations, did not produce different species richness patterns regarding habitat patch size. The exceptions are stressed communities---represented in our models with lowest $\lambda$--- with low dispersal may have a continuous increment of richness with patch size, observed only in random block degradation patterns. This could be related to the optimal patch size for random block habitat loss patterns, which tends to be higher than for regular patterns, so we could hypothesize that the optimal was outside the simulated range for these stressed communities. The aggregation of patches in the random block pattern (implies aggregation of the degraded habitat) could be positive if individual habitat patches are very small because it allows some patches to surpass the minimum size required for survival. For bigger patch sizes it could be detrimental because the aggregation produces even bigger patch sizes that decrease species richness. A similar effect was observed on experimental landscapes [@Fletcher2023], aggregated patterns of habitat loss are detrimental to patch occupancy and eventually could produce the same effects observed in our models about richness.

@@

In previous simulation studies, the importance of fragmentation was demonstrated to be pronounced when the habitat amount is low, approximately around 10% [@Hanski2015; @Rybicki2013]. However, experimental studies have observed that with habitat levels around 10-20%, the impact of fragmentation can be positive or show an intermediate maximum [@Loke2019; @With2016]. When habitat amounts are lower (e.g., 7%), the arrangement of patches appears to have no significant effect. Interestingly, our study replicated similar patterns even with a 40% habitat amount. Moreover, other studies employing comparable habitat amounts (32%) also reported higher species richness with increased fragmentation [@Rybicki2020].

>>>>

---
title: "Why there are more species in several small patches vs few large patches:
  a multi-species modeling approach"
bibliography: sloss_bib.bib
csl: "functional-ecology.csl"
output:
  pdf_document:
    keep_tex: true
    includes:
      in_header: margins.sty
---

@@

2. This work aims to explore spatial multi-species models to verify the SS > SL pattern and determine its causes. We used three types of models: a classical neutral, neutral with habitat selection, and a competitive hierarchical model.  All models have three parameters, namely extinction, reproduction, and dispersal distance. In the classical neutral model, species are equivalent and have the same parameters; when there is a reproduction event they send the propagule to a surrounding patch blindly, and if the patch is already occupied the propagule dies. In the neutral with habitat selection model, the propagules select an empty patch so their survival is ensured. In the hierarchical model, species have competitive dominance and more dominant ones can replace the less dominant with a replacement rate parameter. We use 100 species and start simulating the colonization of an empty area then we destroy a fixed amount of habitat using different regular patch sizes.

@@

    Regular degradation: Habitat is removed in a way that results in habitable patches forming a regular pattern of squares, with side lengths determined by a habitat-size parameter. While the percentage of degraded habitat remains constant, varying the habitat-size parameter leads to either a few large patches (SL) or many smaller ones (SS). The algorithm adjusts either the number or the size of patches to maintain the constant percentage of degradation, especially when only one patch fits within the grid. We quantify this pattern using the mean habitat patch size.

    Random block degradation: Habitat square patches of a fixed size (habitat-size parameter) are located at random throughout the landscape. The sizes are identical to those used for regular degradation simulations, but it results in habitat patches that have a minimum size corresponding to the habitat-size but a maximum size that is bigger because habitat patches can overlap between them producing a larger patch. Then, we characterize this with the mean habitat patch size.

@@

The parameters encompass a range of $growth-rates$ while holding the $death-rate$ fixed at 1. In this model, the relationship between these parameters, $\lambda = growth-rate / death-rate$, is the main determinant of the fate of an individual of a species. In similar non-spatial models, this relationship must exceed 1 for a species to persist. Thus we used a minimum $\lambda$ of 1.7 and fixed the migration rate at 0.0001. We simulated each model variant using the parameters shown in Table 1, with both regular and random block habitat patterns. For each combination of parameters we ran 30 simulations up to 1500 time steps and recorded the number of species, the Shannon Diversity index [@Magurran2013] and the total number of individuals. For some representative combinations of parameters, we take snapshots of the model output showing the typical dynamics on time of the 3 recorded quantities and the rank abundance species distribution [@Newman2005; @May1975]

@@

The spatial pattern of habitat loss is illustrated in Figure 1. Simulations with regular degradation (Figures 1C, 1E, and 1G) produce habitat patches of uniform size, evenly spaced across the landscape. In contrast, simulations using randomly arranged block patterns result in patches of variable sizes, where the habitat-size parameter defines the minimum size, and superposition of patches creates a larger maximum size (Figures 1D, 1F, and 1H). Consequently, the mean patch size is higher for random block degradation than for regular degradation. In some cases, to achieve the targeted percentage of habitat loss, the effective patch size in regular degradation exceeds the defined habitat-size parameter (Figure 2). Simulations with the neutral model illustrate species' spatial distributions before fragmentation (Figures 1A & 1B) and after fragmentation (Figures 1C–F), with habitat destruction depicted in purple. Immediately after fragmentation, species distributions closely resemble pre-fragmentation patterns, but diverge over time.

![Examples of species spatial patterns before and after fragmentation for the neutral model under different habitat loss patterns. Different shades from dark green to white represent distinct species, while destroyed habitat is shown in purple. Simulations were run with a fixed growth rate ($\lambda=2$), a migration rate of 0.0001, and a dispersal distance of 1.5. Panels A & B show spatial patterns before fragmentation. Panels C & D depict patterns immediately after 60% habitat destruction, with patch sizes of 3 and regular degradation (C) or random block degradation (D). Panels E & F illustrate spatial patterns after 1500 time steps, showing population extinction with regular degradation (E). Panels G & H display the spatial pattern after 1500 time steps with a habitat patch size of 29.](Figures/ModelSpatialPatternComparison.png)

In our neutral model simulations with a migration rate of $0.0001$ and regular habitat degradation patterns, species richness varied with habitat patch sizes, influenced by dispersal distance and the $\lambda$ parameter (Figure 2). Species richness increased with $\lambda$, with the highest richness at smaller patch sizes for higher $\lambda$ values. Conversely, a $\lambda$ less than 3 showed reduced richness at the smallest patch sizes for lower dispersal distances. In these cases, only migrants survive in the smallest patches, and no local populations are established (Figures 1E, and S19 & S20).

The interplay of dispersal distances, mean patch sizes, $\lambda$ and percent of habitat destruction impacts community survival and species richness. For $\lambda < 4$ this suggests a threshold patch size influenced by these 3 parameters, for $\lambda = 4$ patch size decreases species richness. The differences between increasing $\lambda$'s almost disappear when dispersal distance is higher. Extended dispersal distances could enhance species survival in scenarios with low $\lambda$ and smaller patches but reduce richness beyond a certain threshold. Similar patterns are evident for Shannon diversity (Figure S1 & S2).

![Relationship between the number of species and mean habitat patch size across varying dispersal distances. The horizontal panels show the relationship between growth rate and death rate, parameter $\lambda$, while the vertical panels represent different percentages of habitat destruction. Simulations were performed using a neutral model with a constant migration rate (0.0001). After 600 time steps, the specified percentage of the habitat was destroyed, making it unavailable for species colonization. The habitat destruction follows a regular pattern, with all patches being of equal size and uniformly spaced from one another.](Figures/Species_MeanPatch_Dispersal_Lambda_pf_regular.png)

![Relationship between the number of species and mean habitat patch size across varying dispersal distances for the neutral model. The horizontal panels show the relationship between growth rate and death rate, parameter $\lambda$, while the vertical panels represent different percentages of habitat destruction. Simulations were performed with a constant migration rate (0.0001). After 600 time steps, the specified percentage of the habitat was destroyed, making it unavailable for species colonization. Habitat blocks of the same size (habitat patch size) were randomly distributed over the simulation grid. As a result, some patches overlapped, leading to a distribution of patch sizes.](Figures/Species_MeanPatch_Dispersal_Lambda_pf_block.png)

For random block habitat degradation the overall pattern is analogous to the regular case (Figure 3), with a peak in richness occurring for the second patch sizes, for $\lambda$ less than 3. However, for values of $\lambda$ greater than 2, maximum richness shifts to the smallest habitat patch size. Patches in the random block simulations are generally larger than in regular degradation, leading to lower species richness and diversity, except at the minimum patch size (Figure 3). Simulations comparing random block and regular habitat patterns (Figure 1) show that, for the smallest patch size, only migrant species persist in regular patterns under neutral and hierarchical models (Figures S19, S20, S27, S28), but not in the neutral model with habitat selection.

![Relationship between the number of species and habitat patch size across panels representing different model versions and varying dispersal distances. The vertical panels correspond to different percentages of habitat destruction. In the neutral model, species are equivalent and share the same parameters; in the hierarchical model, species exhibit competitive dominance, with more dominant species replacing less dominant ones; and in the neutral with habitat selection model, species are neutral but propagules search for empty patches, ensuring survival. Simulations were performed using a fixed growth rate ($\lambda = 2$), with migration and replacement rates of 0.0001 and 0.3, respectively. After 600 time steps, the specified percentage of the habitat was destroyed, making it unavailable for species. The habitat destruction follows a regular pattern, with all patches being of equal size and uniformly spaced from one another.](Figures/AcrossBehav_MeanPatch_Species_Dispersal_Lambda2_pf_regular.png)

![Relationship between the number of species and habitat patch size across panels representing different model versions and varying dispersal distances. The vertical panels correspond to different percentages of habitat destruction. In the neutral model, species are equivalent and share the same parameters; in the hierarchical model, species exhibit competitive dominance, with more dominant species replacing less dominant ones; and in the neutral with habitat selection model, species are neutral but propagules search for empty patches, ensuring survival. Simulations were performed using a fixed growth rate ($\lambda = 2$), with migration and replacement rates of 0.0001 and 0.3, respectively. After 600 time steps, the specified percentage of the habitat was destroyed, making it unavailable for species. Habitat blocks of the same size (habitat patch size) were randomly distributed over the simulation grid. As a result, some patches overlapped, leading to a distribution of patch sizes.](Figures/AcrossBehav_MeanPatch_Species_Dispersal_Lambda2_pf_block.png)

When comparing the three model versions at $\lambda = 2$, the neutral model exhibits higher species richness at larger patch sizes, followed by the neutral with habitat selection model and the hierarchical model (Figures 4 & 5). A similar pattern is seen for Shannon diversity (Figures S11 & S12). At smaller patch sizes, the neutral with habitat selection model supports populations in all cases, whereas in the other models with low dispersal and $\lambda$ below 3, only migrants are present, unable to establish local populations and leading to extinction (Figures 1, S3 & S5). Consequently, intermediate patch sizes with maximum species richness are not observed in the habitat selection model. Additionally, the hierarchical model shows a strong dominance of more competitive species, resulting in very low Shannon diversity at higher dispersal rates (Figures S4 & S6).

More examples of species and habitat spatial patterns are illustrated in Figures S15–S38. In the hierarchical model, differences between simulations with varying dispersal distances are evident. Individual patches are often dominated by a single species, and with short dispersal distances, regions where different species dominate can be established (Figures S23–S30). The impact of minimum patch size on species survival is also clear (Figure 5E, S19, S20, S27, & S28).

The primary difference between varying percentages of habitat destruction is that lower percentages result in a higher number of species in the smallest patch sizes, particularly at higher dispersal distances. In the regular habitat destruction pattern, the increased number of patches shortens the distance between them, facilitating the persistence of local populations. In the random block pattern, patches are generally larger due to overlap, which increases the minimum patch size, allowing local populations to persist as well (Figures 2, 3, 4 & 5).

@@

Our research contributes to the ongoing SLOSS debate in conservation biology, offering new insights based on our model findings. Consistent with previous studies, the results from our simulations primarily support the SS > SL pattern for species retention, as observed across various model scenarios [@Fahrig2020]. This trend underlines the greater efficacy of multiple smaller habitats in preserving species diversity compared to a single large habitat under equal conditions of habitat loss. A critical factor in this dynamic is the balance between reproduction and extinction rates ($\lambda$ parameter) and dispersal relative to the size of the habitat patches, influencing their role as sources of propagules. Notably, the SL > SS scenario emerges only when dispersal limitations prevent patch connectivity, and $\lambda$ is relatively small so patch sizes above a certain threshold are needed to prevent local extinctions, aligning with Rybicki's [-@Rybicki2020] findings. In these cases, an unimodal relationship between species diversity and patch size is observed, with maximum species diversity occurring at larger patch sizes for shorter dispersal distances. Notably, habitat destruction levels do not substantially alter this pattern.

In neutral models, where species are considered equivalent, one species, by chance, tends to dominate a patch [@Hubbell2001]. Consequently, in fragmented landscapes, once a species increases its density, the likelihood of another species replacing it is low. With more patches, the probability of having a greater number of species is heightened. This effect is further amplified when individuals can actively select their habitat, leading to increased survival of propagules. Consequently, fragmentation exhibits a positive influence on the number of species in such scenarios.

@@

Edge effects are present in the neutral and hierarchical models, but not in the neutral with habitat selection model. In the first two models, the edge effect is because individuals in the patch's border have fewer neighbors and the probability of extinction is higher. In the neutral with habitat selection model, the edge effect is not present because the individuals search for empty patches, so the individuals on the border of the patch have the same probability of extinction as the individuals in the interior of the patch. Thus when patches are smaller the edge proportion is higher and the edge effect is more important, but this plays a more important role in the minimal patch size for the survival of species than for the species richness.

The spatial arrangement of habitat patches, contrary to some expectations, did not produce different species richness patterns when habitat patch size is accounted. The exceptions are stressed communities---represented in our models with lowest $\lambda$--- with low dispersal may have a continuous increment of richness with wider increment of patch sizes, observed only in random block degradation patterns.  The aggregation of patches in the random block pattern (implies aggregation of the degraded habitat) could be positive if individual habitat patches are very small because it allows some patches to surpass the minimum size required for survival. For bigger patch sizes it could be detrimental because the aggregation produces even bigger patch sizes that decrease species richness. A similar effect was observed on experimental landscapes [@Fletcher2023], aggregated patterns of habitat loss are detrimental to patch occupancy and eventually could produce the same effects observed in our models about richness.

@@

In previous simulation studies, the importance of fragmentation was demonstrated to be pronounced when the habitat amount is low, approximately around 10% [@Hanski2015; @Rybicki2013]. However, experimental studies have observed that with habitat levels around 10-20%, the impact of fragmentation can be positive or show an intermediate maximum [@Loke2019; @With2016]. When habitat amounts are lower (e.g., 7%), the arrangement of patches appears to have no significant effect. Interestingly, our study replicated similar patterns even within a range of 60 to 20% habitat amount. Moreover, other studies employing comparable habitat amounts (32%) also reported higher species richness with increased fragmentation [@Rybicki2020].
